---
title: "recession predictR"
author: "Wesley Chioh | Jonathan O'Kane"
date: "May 5, 2020"
output: ioslides_presentation
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## introduction

*"Macroeconomic forecasts are often made by separating two objectives: providing the growth rate of the variable of interest (usually GDP) and detecting turning points in the economic cycle."* - [Doz, Ferrara, and Pionnier, 2020](http://econbrowser.com/archives/2020/04/guest-contribution-business-cycle-dynamics-after-the-great-recession-an-extended-markov-switching-dynamic-factor-model)   
  
two models:  
- linear regression: point estimates of future GDP  
- random forest: using gdp and other variables to predict economic state  
  
## data sources  
  
- amt. of electricity generated (monthly) across lower 48 states  
- number of active oil and gas rigs in the US  
- long-term (>= 15weeks) unemployment rate  
- US manufacturing industry confidence indicator  
- yield curve  

## Interactive Plot

```{r eruptions}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  
  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```

## Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## R Output

```{r cars}
summary(cars)
```


