---
title: "recession predictR"
author: "Wesley Chioh | Jonathan O'Kane"
date: "May 5, 2020"
output: ioslides_presentation
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(kableExtra)
library(tidyverse)
library(ggplot2)
```

## introduction

*"Macroeconomic forecasts are often made by separating two objectives: providing the growth rate of the variable of interest (usually GDP) and detecting turning points in the economic cycle."* - [Doz, Ferrara, and Pionnier, 2020](http://econbrowser.com/archives/2020/04/guest-contribution-business-cycle-dynamics-after-the-great-recession-an-extended-markov-switching-dynamic-factor-model)   
  
two models:  
- linear regression: point estimates of future GDP  
- random forest: using gdp and other variables to predict binary economic state  
  
## recession  
  
*"The NBER does not define a recession in terms of two consecutive quarters of decline in real GDP. Rather, a recession is a significant decline in economic activity spread across the economy, lasting more than a few months, normally visible in real GDP, real income, employment, industrial production, and wholesale-retail sales"* - [NBER, 2010](https://www.nber.org/cycles.html)  

the random forest learns which months are contractionary as defined by NBER.  
  
## regression data datasets  
  
 - Probability of Losing a Job  
 - Job Openings per Unemployed Person  
 - Ratio of Hires to Openings  
 - Business Survey  
 - Yield Curve  
 - Economic Policy Index  
 - Equity Uncertainty  
 - Unemployment Rate (YoY)
  
## random forest datsets   
  
- gdp  
- amt. of electricity generated (monthly) across lower 48 states  
- number of active oil and gas rigs in the US  
- long-term (>= 15weeks) unemployment rate  
- US manufacturing industry confidence indicator  
- yield curve  
  
## data sources  
  
- Energy Information Administration  
- Federal Reserve  
- Organization for Economic Co-operation and Development  
- National Bureau of Economic Research  
- IHS Markit  


## annualized gdp change forecast  
looking ahead based on February 2020 data:    
  
```{r forecast df}
month <- c("March 2020", "April 2020","May 2020")
one_month <- c("6.27%", NA, NA)
two_month <- c("1.4%", "5.0%", NA)
three_month <- c("2.62%", "2.02%", "3.37%")
forecast_df <- cbind(month, one_month, two_month, three_month) %>%
  data.frame()
colnames(forecast_df) <- c("Month", "One Month", "Two Months", "Three Months")
kable(forecast_df) %>%
  kable_styling(bootstrap_options = c("striped","hover", "condensed"))
```


## shiny x gdp forecasting

```{r regression}
inputPanel(
  selectInput("n_lags", label = "Number of Months:",
              choices = c(1,2,3,4,5,6), selected = 1),
  sliderInput("train_test_adjust", label = "Train-Test Split:",
              min = 0.1, max = 0.9, value = 0.7, step = 0.1)
)

lagged_predictors <- function(n_lags, predictor_df, n_predictors=10){
  pred_df <- sapply(1:n_predictors, function(x){
    lag(predictor_df[,x],n_lags)
  } )
  pred_df <- as.data.frame(pred_df)
  return(pred_df)
}
predictors <- readRDS("./models/predictors.RDS")

renderPlot({
  # calculating lag
  predictors_oml <- lagged_predictors(as.numeric(input$n_lags),predictors)
  predictors_oml <- cbind(predictors$GDP_LOG, predictors_oml) %>%
    drop_na()
  colnames(predictors_oml)[1] <- "GDP_LOG"

  # sampling then testing
  testIDs <- sample(1:dim(predictors_oml)[1], input$train_test_adjust*dim(predictors_oml)[1])
  training_predictors_oml <- predictors_oml[testIDs,]
  testing_predictors_oml <- predictors_oml[-testIDs,]
  gdp_oml_lm <- lm(GDP_LOG ~., data = training_predictors_oml)

  pred_gdp_log_oml <- predict.lm(gdp_oml_lm, newdata = testing_predictors_oml)
  ggplot() + 
    geom_point(aes(x = testing_predictors_oml$GDP_LOG, y = pred_gdp_log_oml)) + 
    geom_abline(intercept = 0, slope = 1) + 
    labs(x = "Actual Logged GDP",
         y = "Predicted Logged GDP")
})
```

## dull side of things  
```{r eval = FALSE, echo=TRUE}
lagged_predictors <- function(n_lags, predictor_df, n_predictors=10){
  pred_df <- sapply(1:n_predictors, function(x){
    lag(predictor_df[,x],n_lags) })
  pred_df <- as.data.frame(pred_df)
  return(pred_df)}
predictors <- readRDS("./models/predictors.RDS")
predictors_oml <- lagged_predictors(as.numeric(input$n_lags),predictors)
predictors_oml <- cbind(predictors$GDP_LOG, predictors_oml) %>% drop_na()
testIDs <- sample(1:dim(predictors_oml)[1], 
                  input$train_test_adjust*dim(predictors_oml)[1])
training_predictors_oml <- predictors_oml[testIDs,]
testing_predictors_oml <- predictors_oml[-testIDs,]
gdp_oml_lm <- lm(GDP_LOG ~., data = training_predictors_oml)
pred_gdp_log_oml <- predict.lm(gdp_oml_lm, 
                               newdata = testing_predictors_oml)
```


## growing the amazon  
  
```{r eval = FALSE, echo = TRUE}
trctrl <- trainControl(method = "repeatedcv", 
                       number = 10, repeats = 5, search = "grid")
mtry <- ncol(train_x)
ntrees <- 101
tunegrid <- expand.grid(.mtry = c(2:mtry))
metric <- "Accuracy"
cl <- makePSOCKcluster(5)
registerDoParallel(cl)
rf_recession <- train(x = train_x, y = train_y, method = "rf", 
                      metric = metric, trControl = trctrl, 
                      tuneGrid = tunegrid, ntree = ntrees)
stopCluster(cl)
```

## beneath the canopy  
left-split <= s; right-split >s where s is split 
```{r, warning=FALSE, message=FALSE}
tree_plot <- readRDS("./models/tree_plot.RDS")
# if split is k, LHS is <= k,RHS is >k
tree_plot
```


## results
with a **70-30** training-test split:  
  
```{r confusion matrix}
rf_cm <- readRDS("./models/confusion_matrix_1mth_all.RDS")
kable(rf_cm$table, caption = "confusion matrix | pred (v), ref (h)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = T, position = "left")
kable(cbind(t(rf_cm$byClass[5:8]),t(rf_cm$overall[1])), caption = "random forest performance") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F, position = "left")
```

## probabilistically speaking

```{r random forest plot}
rf_prob_plot <- readRDS("./models/ggplot_prob_model.RDS")
rf_prob_plot
```

## pushing it just a little farther

```{r}
rf_cm <- readRDS("./models/confusion_matrix_2mth_all.RDS")
kable(rf_cm$table, caption = "confusion matrix | pred (v), ref (h)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = T, position = "left")
kable(cbind(t(rf_cm$byClass[5:8]),t(rf_cm$overall[1])), caption = "random forest performance") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F, position = "left")
```

## probabilistically speaking again

```{r}
rf_prob_plot <- readRDS("./models/ggplot_prob_model_2.RDS")
rf_prob_plot
```


## questions  

source code: github.com/westerleyy/recession_predictR
