---
title: "predictR"
author: "Wesley Chioh"
date: "April 19, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# setting up shop 
library(tidyverse)
library(ggplot2)
library(rjson)
library(jsonlite)
library(eia)
library(ggfortify)
library(fredr)
library(ggpmisc)
library(zoo)
library(plotly)
library(caret)

# load API key
eia_api_key <- readtext::readtext("eia_api_key.txt") %>%
  .$text
fred_api_key <- readtext::readtext("fred_api_key.txt") %>%
  .$text
eia_set_key(eia_api_key)
fredr_set_key(fred_api_key)

# load datasets
electricity <- read_csv("monthly_electricity.csv")
rigs <- read_csv("monthly_rigs_count.csv")
gdp <- readxl::read_excel("Monthly GDP for Project.xlsx")
business_sentiment <- read_csv("business_tendencies.csv")
```

## Data Processing  

Relevant time scale: Monthly  
Datasets: Electricity generation, Rigs count, GDP, OECD Business Tendency Survey  
  
At first, I used the GDP data to determine whether there is a recession that particular quarter or not. But I realized that it is better to use the NBER-defined recession months. So that is what I did. Use the HTML table to compare between the self-defined recession months and NBER-defined ones. 
```{r data wrangling}
gdp_complete <- gdp %>%
  mutate(date = as.yearmon(substr((Date), 1, 7)),
         `GDP (Continuously Compounding)` = `GDP (Continuously Compounding)`*100,
         `GDP (YoY % Change)` = `GDP (YoY % Change)`*100) %>%
  filter(date >= "Jan 1973") %>%
  select(date, `GDP (YoY % Change)`, `GDP (Continuously Compounding)`)
electricity_complete <- electricity %>%
  mutate(date = as.yearmon(date),
         rate_generated_change = (value - lag(value,1))/lag(value,1),
         rate_generated_change_secondorder = (rate_generated_change - lag(rate_generated_change,1))/lag(rate_generated_change,1)) %>%
  filter(date <= "Nov 2019") %>%
  select(date, rate_generated_change, rate_generated_change_secondorder) 
rigs_complete <- rigs %>%
  mutate(date = as.yearmon(substr(date, 1, 7))) %>%
  select(date, change_first_order, change_second_order) %>%
  filter(date <= "Nov 2019")
colnames(rigs_complete)[2:3] <- c("rigs_change_first_order", "rigs_change_second_order")
business_sentiment <- business_sentiment %>%
  mutate(date = as.yearmon(substr(date, 1, 7)),
         sentiment_change_first_order = (value - lag(value,1))/lag(value,1)) %>%
  select(date, value, distance_peak, sentiment_change_first_order)
colnames(business_sentiment)[2] <- "business_sentiment"
sentiment_period <- rep(0, dim(business_sentiment)[1])
for (i in 2:dim(business_sentiment)[1]){
  sentiment_i = business_sentiment$sentiment_change_first_order[i]
  #print(sentiment_i)
  sentiment_h = business_sentiment$sentiment_change_first_order[i-1]
  #print(sentiment_h)
  sentiment_h[is.na(sentiment_h)] <- 0
  if(sign(sentiment_i) == sign(sentiment_h)){
    sentiment_period[i] = sentiment_period[i-1]+1
  }
  else{
    sentiment_period[i] = 1
  }
}
business_sentiment$sentiment_period <- sentiment_period


# inner join all together and to get recession dates
econ_situation_nber_df <- inner_join(gdp_complete, electricity_complete, by = "date")
econ_situation_nber_df <- inner_join(econ_situation_nber_df, rigs_complete, by = "date")
econ_situation_nber_df <- inner_join(econ_situation_nber_df, business_sentiment, by = "date")
NBER_recession_months <- econ_situation_nber_df$date %>%
  .[c(11:27, 85:90, 103:119, 211:219, 339:347, 420:438)]
econ_situation_nber_df <- econ_situation_nber_df %>%
  mutate(NBER_recession = ifelse(date %in% NBER_recession_months, "Y", "N"),
         NBER_recession_next = as.factor(lead(NBER_recession,1))) %>%
  as.data.frame() %>%
  drop_na()

DT::datatable(econ_situation_nber_df)
```

## Random Forest

So I created a random forest model and asked it a very simple question: Will there be a recession the following month or not, given all the data?  
The model was given the data on change in electricity generation, and change in number of oil and natural gas rigs in the US.  
The model was trained and tested on a 70:30 randomly sampled split. 
  
How to read a confusion matrix? The diagonal running from top-left to bottom-right is the number of correct predictions. No Information Rate means that if the model just predicted `Y` all the time, it would get it right how many % of the time. So clearly, this random forest model is better.  

```{r random_forest 2}
set.seed(2507)
# creating test-train data
testIDs <- createDataPartition(econ_situation_nber_df$NBER_recession_next, p = 0.7, list = F)
train_x <- econ_situation_nber_df[testIDs,c(4:11)]
train_y <- econ_situation_nber_df$NBER_recession_next[testIDs]
test_x <- econ_situation_nber_df[-testIDs, c(4:11)]
test_y <- econ_situation_nber_df$NBER_recession_next[-testIDs] 

# model tuning 
trctrl <- trainControl(method = "repeatedcv", 
                       number = 10, repeats = 5, search = "grid")
mtry <- ncol(train_x)
ntrees <- 71
tunegrid <- expand.grid(.mtry = c(2:mtry))
metric = "Accuracy"
# random forest model 
rf_recession <- train(x = train_x, y = train_y, method = "rf", metric = metric, trControl = trctrl, tuneGrid = tunegrid)
plot(rf_recession)
predict_y <- predict(rf_recession, newdata = test_x)
confusionMatrix(predict_y, reference = test_y)

```

