---
title: "predictR"
author: "Wesley Chioh"
date: "April 19, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# setting up shop 
library(tidyverse)
library(ggplot2)
library(eia)
library(fredr)
library(zoo)
library(plotly)
library(caret)
library(gbm)
library(doParallel)
library(stargazer)
library(rjson)
library(lubridate)
library(jsonlite)
library(ggraph)
library(igraph)
set.seed(1728)

# load API key
eia_api_key <- readtext::readtext("eia_api_key.txt") %>%
  .$text
fred_api_key <- readtext::readtext("fred_api_key.txt") %>%
  .$text
eia_set_key(eia_api_key)
fredr_set_key(fred_api_key)

# load datasets
gdp <- readxl::read_excel("Monthly GDP for Project.xlsx")

```

## Data Processing  

Relevant time scale: Monthly  
Datasets: Electricity generation, Rigs count, GDP, OECD Business Tendency Survey, Percentage of Civilian Workforce Unemployed >15 weeks

```{r downloading datasets and initial processing, warning=FALSE}
# load API key
eia_api_key <- readtext::readtext("eia_api_key.txt") %>%
  .$text
fred_api_key <- readtext::readtext("fred_api_key.txt") %>%
  .$text
eia_set_key(eia_api_key)
fredr_set_key(fred_api_key)

## pct labor force unemloyed for more than 15 weeks
LT_unemployed <- fredr(series_id = "U1RATE", observation_start = as.Date("1973-01-01"))
## yield curve 10year minue 3month
yield_curve <- fredr(series_id = "T10Y3M", observation_start=as.Date('1947-01-01'))

## business tendenc
business_tendenc <- fredr(series_id = "BSCICP03USM665S", observation_start = as.Date("1973-01-01"))

## finding local peaks where m is the number of points on either side of the peak 
find_peaks <- function (x, m = 3){
  shape <- diff(sign(diff(x, na.pad = FALSE)))
  pks <- sapply(which(shape < 0), FUN = function(i){
    z <- i - m + 1
    z <- ifelse(z > 0, z, 1)
    w <- i + m + 1
    w <- ifelse(w < length(x), w, length(x))
    if(all(x[c(z : i, (i + 2) : w)] <= x[i + 1])) return(i + 1) else return(numeric(0))
  })
  pks <- unlist(pks)
  pks
}

distance_from_peak <- function(current_position, peaks_vec, time_series_value){
  if (current_position > peaks_vec[1]){
  distance <- peaks_vec-current_position
  closest <- min(distance)
  distance_pct <- (time_series_value[current_position] - time_series_value[current_position+closest])/time_series_value[current_position+closest]
  distance_pct <- ifelse(is.null(distance_pct) == T, 0, distance_pct)
  distance_pct
  } else { distance_pct <- 0
  distance_pct}
}

business_local_peaks <- find_peaks(business_tendenc$value)
distance_peak <- sapply(1:dim(business_tendenc)[1], function(x){
  distance_pct <- distance_from_peak(x, business_local_peaks, business_tendenc$value)
  distance_pct
})

business_tendencies <- cbind(business_tendenc, distance_peak)
business_tendencies[is.na(business_tendencies)] <- 0
write.csv(business_tendencies, "./data/business_tendencies.csv")
business_sentiment <- business_tendencies

## Electricity Data
# hourly demand

lower48_hourly_demand <- paste("http://api.eia.gov/series/?series_id=EBA.US48-ALL.D.H&api_key=", eia_api_key, sep = "") %>%
  fromJSON() 
lower48_hourly_demand_list <- lower48_hourly_demand$series %>%
  select(data)%>%
  .[[1]] 
date_time <- lower48_hourly_demand_list[[1]][,1]
demand <- lower48_hourly_demand_list[[1]][,2]
lower48_hourly_demand_df <-  cbind(date_time, demand) %>%
  data.frame(stringsAsFactors = F) %>%
  mutate(date = as_date(ymd(substr(date_time, 1, 8))),
         hour = as.numeric(substr(date_time, 10, 11)))

## all generation
# last updated 4/26
all_electricity_generation_monthly <- eia_series(id = "TOTAL.ELETPUS.M")
all_electricity_generation_monthly_df <- unnest(all_electricity_generation_monthly, cols="data")
all_electricity_generation_monthly_df <- all_electricity_generation_monthly_df %>%
  select(series_id, units, value, date, year, month)

all_electricity_generation_monthly_df <- all_electricity_generation_monthly_df %>%
  select(series_id, value, date, year, month) %>%
  mutate(terawatt_value = value/1000, 
         units = "Terawatthours",
         date = as.yearmon(unique(substr(date, 1,7))))

# hourly generation
lower48_hourly_generation <- eia_series(id = "EBA.US48-ALL.NG.H")
lower48_hourly_generation_df <- unnest(lower48_hourly_generation, cols = "data") %>%
  select(series_id, units, value, date, year, month)

# monthly generation from hourly

lower48_monthly_generation_df <- lower48_hourly_generation_df %>%
  group_by(year, month) %>%
  summarise(value = (sum(value))/1000000,
            units = "Terawatthours",
            date = unique(substr(date, 1,7)))
lower48_monthly_generation_df$date <- as.yearmon(lower48_monthly_generation_df$date)

# intersecting all states with lower 48 | then calculate prop lower48 to extrapolate retroactively
intersecting_months <- inner_join(lower48_monthly_generation_df, all_electricity_generation_monthly_df, by = "date")
colnames(intersecting_months)[3] <- "lower48_terawatt" 
intersecting_months$proportion <- intersecting_months$lower48_terawatt/intersecting_months$terawatt_value
lower48_prop <- fivenum(intersecting_months$proportion)[3]
all_electricity_generation_monthly_extrapolated_df <- all_electricity_generation_monthly_df %>%
  mutate(lower48_terawatt = terawatt_value*lower48_prop) %>%
  filter(date < "Jul 2015") %>%
  select(year, month, lower48_terawatt, units, date)
colnames(all_electricity_generation_monthly_extrapolated_df)[3] <- "value"
lower48_monthly_generation_extrapolated_df <- bind_rows(all_electricity_generation_monthly_extrapolated_df, lower48_monthly_generation_df)
write.csv(lower48_monthly_generation_extrapolated_df, "./data/monthly_electricity.csv")
electricity <- lower48_monthly_generation_extrapolated_df

## crude oil and nat gas rigs in operation 
total_rigs <- eia_series(id = "TOTAL.OGNRPUS.M")
total_rigs_df <- unnest(total_rigs, cols = "data") %>%
  select(series_id, units, value, date, year, month)
total_rigs_df <- total_rigs_df %>%
  arrange(date) %>%
  mutate(change_first_order = round((value - lag(value, 1))/lag(value,1),4),
         change_second_order = ifelse(lag(change_first_order,1) == 0, 0, (change_first_order - lag(change_first_order, 1))/lag(change_first_order,1)))
write.csv(total_rigs_df, "./data/monthly_rigs_count.csv")
rigs <- total_rigs_df
```
  

```{r data wrangling}
gdp_complete <- gdp %>%
  mutate(date = as.yearmon(substr((Date), 1, 7)),
         `GDP (Continuously Compounding)` = `GDP (Continuously Compounding)`*100,
         `GDP (YoY % Change)` = `GDP (YoY % Change)`*100, 
         gdp_change_lagged = lag(`GDP (Continuously Compounding)`, 1)) %>%
  filter(date >= "Jan 1973") %>%
  select(date, `GDP (YoY % Change)`, `GDP (Continuously Compounding)`, gdp_change_lagged)
electricity_complete <- electricity %>%
  mutate(date = as.yearmon(date),
         rate_generated_change = (value - lag(value,1))/lag(value,1),
         rate_generated_change_secondorder = (rate_generated_change - lag(rate_generated_change,1))/lag(rate_generated_change,1)) %>%
  select(date, rate_generated_change, rate_generated_change_secondorder) 
rigs_complete <- rigs %>%
  mutate(date = as.yearmon(substr(date, 1, 7))) %>%
  select(date, change_first_order, change_second_order)
colnames(rigs_complete)[2:3] <- c("rigs_change_first_order", "rigs_change_second_order")
business_sentiment <- business_sentiment %>%
  mutate(date = as.yearmon(substr(date, 1, 7)),
         sentiment_change_first_order = (value - lag(value,1))/lag(value,1)) %>%
  select(date, value, distance_peak, sentiment_change_first_order)
colnames(business_sentiment)[2] <- "business_sentiment"
sentiment_period <- rep(0, dim(business_sentiment)[1])
for (i in 2:dim(business_sentiment)[1]){
  sentiment_i = business_sentiment$sentiment_change_first_order[i]
  #print(sentiment_i)
  sentiment_h = business_sentiment$sentiment_change_first_order[i-1]
  #print(sentiment_h)
  sentiment_h[is.na(sentiment_h)] <- 0
  if(sign(sentiment_i) == sign(sentiment_h)){
    sentiment_period[i] = sentiment_period[i-1]+1
  }
  else{
    sentiment_period[i] = 1
  }
}
business_sentiment$sentiment_period <- sentiment_period

LT_unemployed_complete <- LT_unemployed %>%
  mutate(date = as.yearmon(substr(date, 1, 7)),
         rate_change_first_order = (value - lag(value,1))/lag(value,1)) %>%
  select(date, value, rate_change_first_order)
colnames(LT_unemployed_complete)[2] <- "LT_unemployment"
rate_change_period <- rep(0, dim(LT_unemployed_complete)[1])
threshold_period <- rep(0, dim(LT_unemployed_complete)[1])
for (i in 2:dim(LT_unemployed_complete)[1]){
  sentiment_i = LT_unemployed_complete$rate_change_first_order[i]
  #print(sentiment_i)
  sentiment_h = LT_unemployed_complete$rate_change_first_order[i-1]
  #print(sentiment_h)
  sentiment_h[is.na(sentiment_h)] <- 0
  if(sign(sentiment_i) == sign(sentiment_h)|sign(sentiment_i)==0){
    rate_change_period[i] = rate_change_period[i-1]+1
  }
  else{
   rate_change_period[i] = 1
  }
  if(LT_unemployed_complete$LT_unemployment[i] <= 1.5){
    threshold_period[i] = threshold_period[i-1]+1
  }
}
LT_unemployed_complete$threshold_period <- threshold_period
LT_unemployed_complete$rate_change_period <- rate_change_period

yield_curve[is.na(yield_curve)] <- 0
yield_curve <- yield_curve %>%
  drop_na() %>%
  mutate(date = as.yearmon(substr(date, 1, 7)),
         value = ifelse(value < 0, -1, 1)) %>%
  group_by(date) %>%
  summarise(value = min(value)) %>%
  mutate(value = as.factor(value))
colnames(yield_curve)[2] <- "curve"

# inner join all together and to get recession dates
econ_situation_nber_df <- inner_join(gdp_complete, electricity_complete, by = "date")
econ_situation_nber_df <- inner_join(econ_situation_nber_df, rigs_complete, by = "date")
econ_situation_nber_df <- inner_join(econ_situation_nber_df, business_sentiment, by = "date")
econ_situation_nber_df <- inner_join(econ_situation_nber_df, LT_unemployed_complete, by = "date")
NBER_recession_months <- econ_situation_nber_df$date %>%
  .[c(11:27, 85:90, 103:119, 211:219, 339:347, 420:438)]
econ_situation_nber_df <- inner_join(econ_situation_nber_df, yield_curve, by = "date")
econ_situation_nber_df <- econ_situation_nber_df %>%
  mutate(NBER_recession = ifelse(date %in% NBER_recession_months, "Y", "N"),
         NBER_recession_next = as.factor(lead(NBER_recession,1)),
         NBER_recession_two = as.factor(lead(NBER_recession,2))
         ) %>%
  as.data.frame() %>%
  drop_na()

DT::datatable(econ_situation_nber_df)
```

## Random Forest Modeling  
  
Two random forest models were trained. One to predict the probability of a recession the next month, and another for the following month. A previously-trained model was loaded to save time when knitting this document. 

```{r random forest onemonth}
# creating test-train data
testIDs <- createDataPartition(econ_situation_nber_df$NBER_recession_next, p = 0.7, list = F)
train_x <- econ_situation_nber_df[testIDs,c(4:17)]
train_y <- econ_situation_nber_df$NBER_recession_next[testIDs]
test_x <- econ_situation_nber_df[-testIDs, c(4:17)]
test_y <- econ_situation_nber_df$NBER_recession_next[-testIDs] 

# model tuning 
trctrl <- trainControl(method = "repeatedcv", 
                       number = 10, repeats = 5, search = "grid")
mtry <- ncol(train_x)
ntrees <- 101
tunegrid <- expand.grid(.mtry = c(2:mtry))
metric = "Accuracy"
# following can be commented out if loading a pretrained model from file 
# parallel processing to speed things up
# random forest model 
# cl <- makePSOCKcluster(5)
# registerDoParallel(cl)
# rf_recession <- train(x = train_x, y = train_y, method = "rf", metric = metric, trControl = trctrl, tuneGrid = tunegrid, ntree = ntrees)
# stopCluster(cl)
# saveRDS(rf_recession, "./models/rf_1mth_all.RDS")
rf_recession <- readRDS("./models/rf_1mth_all.RDS")
predict_y <- predict(rf_recession, newdata = test_x)
predict_y_prob <- predict(rf_recession,newdata = test_x, "prob")
predict_y_prob$date <- econ_situation_nber_df$date[-testIDs]
predict_y_prob$actual <- test_y
predict_y_prob$predicted <- predict_y
cm_1mth_all <- confusionMatrix(predict_y, reference = test_y)
saveRDS(cm_1mth_all, "./models/confusion_matrix_1mth_all.RDS")
cm_1mth_all$overall

prob_model <- ggplotly(ggplot(predict_y_prob, aes(x = date, y = Y)) +
                         geom_point(aes(color = actual)) + 
                         geom_hline(yintercept = 0.50) + 
                         labs(y = "Probability", x = "Date", colour = "NBER", 
                              title = "Probability of a Recession a Month Ahead", subtitle = "Probabilistic Prediction of a Random Forest Model"))
saveRDS(prob_model, "./models/ggplot_prob_model.RDS")
prob_model
```



```{r random_forest two month, fig.height=8, fig.width=8}
# creating test-train data
testIDs <- createDataPartition(econ_situation_nber_df$NBER_recession_two, p = 0.7, list = F)
train_x <- econ_situation_nber_df[testIDs,c(4:17)]
train_y <- econ_situation_nber_df$NBER_recession_two[testIDs]
test_x <- econ_situation_nber_df[-testIDs, c(4:17)]
test_y <- econ_situation_nber_df$NBER_recession_two[-testIDs] 

# model tuning 
trctrl <- trainControl(method = "repeatedcv", 
                       number = 10, repeats = 5, search = "grid")
mtry <- ncol(train_x)
ntrees <- 101
tunegrid <- expand.grid(.mtry = c(2:mtry))
metric = "Accuracy"
# following can be commented out if loading a pretrained model from file 
# parallel processing to speed things up
# random forest model 
# cl <- makePSOCKcluster(5)
# registerDoParallel(cl)
# rf_recession <- train(x = train_x, y = train_y, method = "rf", metric = metric, trControl = trctrl, tuneGrid = tunegrid, ntree = ntrees)
# stopCluster(cl)
# saveRDS(rf_recession, "./models/rf_2mth_all.RDS")
rf_recession <- readRDS("./models/rf_2mth_all.RDS")
predict_y <- predict(rf_recession, newdata = test_x)
predict_y_prob <- predict(rf_recession,newdata = test_x, "prob")
predict_y_prob$date <- econ_situation_nber_df$date[-testIDs]
predict_y_prob$actual <- test_y
predict_y_prob$predicted <- predict_y
cm_2mth_all <- confusionMatrix(predict_y, reference = test_y)
saveRDS(cm_2mth_all, "./models/confusion_matrix_2mth_all.RDS")
cm_2mth_all$byClass

prob_model_2 <- ggplotly(ggplot(predict_y_prob, aes(x = date, y = Y)) +
                         geom_point(aes(color = actual)) + 
                         geom_hline(yintercept = 0.50) + 
                         labs(y = "Probability", x = "Date", colour = "NBER", 
                              title = "Probability of a Recession 2 Months Ahead", subtitle = "Probabilistic Prediction of a Random Forest Model"))
saveRDS(prob_model_2, "./models/ggplot_prob_model_2.RDS")
prob_model_2
```

### Visualizing Random Forest  
  
The following selects a decision tree from the random forest model and plots it. `tree_func` adopted wholesale from [here](https://shiring.github.io/machine_learning/2017/03/16/rf_plot_ggraph)

```{r plotting random forest, warning=FALSE, message=FALSE}
# https://shiring.github.io/machine_learning/2017/03/16/rf_plot_ggraph

tree_func <- function(final_model, 
                      tree_num) {
  
  # get tree by index
  tree <- randomForest::getTree(final_model, 
                                k = tree_num, 
                                labelVar = TRUE) %>%
    tibble::rownames_to_column() %>%
    # make leaf split points to NA, so the 0s won't get plotted
    mutate(`split point` = ifelse(is.na(prediction), `split point`, NA))
  
  # prepare data frame for graph
  graph_frame <- data.frame(from = rep(tree$rowname, 2),
                            to = c(tree$`left daughter`, tree$`right daughter`))
  
  # convert to graph and delete the last node that we don't want to plot
  graph <- graph_from_data_frame(graph_frame) %>%
    delete_vertices("0")
  
  # set node labels
  V(graph)$node_label <- gsub("_", " ", as.character(tree$`split var`))
  V(graph)$leaf_label <- as.character(tree$prediction)
  V(graph)$split <- as.character(round(tree$`split point`, digits = 2))
  
  # plot
  plot <- ggraph(graph, 'dendrogram') + 
    theme_bw() +
    geom_edge_link() +
    geom_node_point() +
    geom_node_text(aes(label = node_label), na.rm = TRUE, repel = TRUE) +
    geom_node_label(aes(label = split), vjust = 2.5, na.rm = TRUE, fill = "white") +
    geom_node_label(aes(label = leaf_label, fill = leaf_label), na.rm = TRUE, 
					repel = TRUE, colour = "white", fontface = "bold", show.legend = FALSE) +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_blank(),
          panel.background = element_blank(),
          plot.background = element_rect(fill = "white"),
          panel.border = element_blank(),
          axis.line = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          plot.title = element_text(size = 18))
  
  print(plot)
}

tree_num <- which(rf_recession$finalModel$forest$ndbigtree == min(rf_recession$finalModel$forest$ndbigtree))
tree_plot <- tree_func(final_model = rf_recession$finalModel, 12)
saveRDS(tree_plot, "./models/tree_plot.RDS")
```

